\text{Additive drift with tail bounds}

### 2021-09-26 article.tex
fix english

\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true}
\definecolor{darkGreen}{RGB}{31, 140, 47}
\definecolor{darkBlue}{RGB}{31, 53, 125}   
\title{\text{Additive drift with tail bounds}}

@@

\end{proof}
\subsection{Theorem for sub-gaussian processes}
\cm
\subsection{Comparing requirements}
\label{sec:compare}
To use some versions of drift theorems, we need some requirements on the sequence of random variables to be satisfied, such as the following:
$$
    \exists \delta \geq 0,\ c \in \mathbb{R} \ \forall  t \geq 0 \ \forall \gamma \in [0, \delta] $$$$
    E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq e^{\frac{c \gamma^2}{2}},$$

оr

$$\exists \delta \geq 0,\ p \geq 1 \ \forall t \geq 0 $$$$
    E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq 1 - \frac{1}{p}\ .$$

To compare rigors of requirements we need to explore behavior $E[e^{\gamma X}]$ regarding to $\gamma$. Of course this behavior depends on distribution type, hence consider common facts.

@@

Consider the definition of expectation of a function
\[
    E[e^{\gamma X}] = \int_{-\infty}^{+\infty} f(x) e^{\gamma x} \,dx,
\]

where $f(x)$ is a probability density function.

@@

Let's take the first derivative $\gamma$

@@

Note that $\forall x \leq 0 e^{\gamma x} \leq 1$ and $\forall x \geq 0 e^{\gamma x} \geq 1 + \gamma x$. Hence because $\forall x f(x) \geq 0$

@@

And then

@@

Denote $\int_{0}^{+\infty} x^2 f(x) \, dx = c \geq 0$ and then if we reintegrate back, we get a restriction on the expectation from below

@@

In the same way consider expectation of function by definition

\begin{align*}
    E[e^{\gamma X}] = \sum_{i = 1}^{+\infty} e^{\gamma x_i} \Pr[x = x_i],
\end{align*}

where $\Pr$ is a probability function.

@@

After taking the first derivative we may split terms from $P$ set with other terms
% В таком случае, аналогично возьмем производную и рассмотрим пот отдельности слагаемые из множества P и остальные:

@@

Again denote $c = \sum_{i \in P} x_i^2 Pr[x = x_i] \geq 0$ and obtain a result similar to the previous paragraph

@@

Because formulas for both cases are same all further conclusions will be same too.

@@

In this case, the expectation of the exponent cannot be lower than 1, and also in the vicinity of zero it grows faster than any$e^{\frac{\gamma^2 c}{2}}$, because

@@

But $g(0) = 0$, while $E[X] > 0$, therefore, that in any vicinity of zero

@@

Hence both of requirements are not feasible.

@@

In this section we will consider some specific family of continuos distributions with symmetric probability density functions, because their expectation equals 0.

@@

\begin{align*}
    \infint{f(x)} & = \infint{c\expx{a}} = 1
\end{align*}

@@

Let's make a replacement

@@

then

@@

Let's find n-th momentum of $X$ (expectation от $X^n$)

@@

\begin{align*}
    E[e^{\gamma X}] = E\left[\sum_{n = 0}^{+\infty}\frac{(\gamma X)^n}{n!}\right] & = \sum_{n = 0}^{+\infty}\gamma^n\frac{E[X^n]}{n!} = \sum_{n = 0}^{+\infty}\gamma^{2n}\frac{E[X^{2n}]}{2n!} =                                \\
    & = \sum_{n = 0}^{+\infty}\frac{(\gamma^2)^{n}}{n!}\left(\frac{E[X^{2n}]n!}{2n!}\right) = \sum_{n = 0}^{+\infty}\frac{(\gamma^2)^{n}}{n!} a_n
\end{align*}

@@

Consider sequence $\{\sqrt[n]{a_n}\}$

@@

Let's take a look on convergence of previous sequence regarding to values of the $a$
\begin{caseof}
    \case{$a \geq 2$}{

        Then $A$ exist, which means existing $C > 0$ with property
        $\forall n \geq 0 a_n \leq C^n$.
        Hence $E[e^{\gamma X}] \leq e^{\gamma^2 C}.$
    }
    \case{$a < 2$}{

@@

We can compute same limit for $\{\sqrt[n]{\frac{a_n}{n!}}\}$ and threshold regarding to $a$ is 1, where only for $a \geq 1$ series converge.

@@

For $a \in (1, 2)$ obviously that asymptotics is greater that square root. Hence sub-gausality is not feasible.

@@

So both requirements are feasible since

@@

So proving that the process is sub-gaussian is tantamount to finding $p$ in the second requirements.

@@

If $E[X] > 0$ then in some neighborhood around $\gamma = 0$ would be greater than any $e^{\frac{c \gamma^2}{2}}$, because if

@@

Example

\begin{figure}[h!]
    \begin{center}
     \begin{tikzpicture}[scale = 0.8]
        \begin{axis}[
            axis lines = left,
            xlabel = \(\gamma\),
            ylabel = {\(f(\gamma)\)},
            xmin=0, xmax=0.4,
            ymin=0.95, ymax=1.5
        ]
        % \addplot[color=darkGreen,
        %     domain=0:0.9,
        %     dashed,
        %     line width=1pt
        % ]{1 + (1) * x + (1.925) / 2 * x * x};
        % \addlegendentry{\(B_X(\gamma)\)}
        \addplot[color=green,
            domain=0.04:0.3,
            line width=0.8pt
        ]{exp((50 * x^2) / 2)};

        \addplot[color=red,
            domain=0:0.6,
            line width=0.8pt,
            dashed
        ]{exp((4 * x^2) / 2)};

        \addplot[color=green,
            domain=0.1:0.6,
            line width=0.8pt
        ]{exp((20 * x^2) / 2)};

        % \addplot[color=red,
        %     domain=0:0.6,
        %     line width=1.2pt
        % ]{exp((15 * x^2) / 2)};

        \addplot[color=red,
            domain=0:0.6,
            line width=0.8pt,
            dashed
        ]{exp((20 * x^2) / 2)};

        \addplot[color=red,
            domain=0:0.3,
            line width=0.8pt,
            dashed
        ]{exp((50 * x^2) / 2)};

@@

        % \addplot[color=black,mark=*] coordinates {(0.3578, 0.9174)};
        % \addplot[color=black,mark=*] coordinates {(0.15, 0.9631)};
        \addplot[color=red,
            domain=0:3,
            dashed
        ]{1};
        \end{axis}
     \end{tikzpicture}
    \end{center}
    % \caption{Plot of the function $Ee_X(\gamma)$ with its bound $B_X(\gamma)$, where $X$ has probability density function $f(x) = 0.65 e^{-0.65(x + 2)}$.}
    % \label{fig:negativeDriftPic2}
\end{figure}

\subsection*{Заключение}

In the case of $\textbf{\text{zero}}$ expectation we cannot always guarantee subgausality, which has been proved by the example of series of even functions $\expx{a}$.

>>>>

\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true}
\definecolor{darkGreen}{RGB}{31, 140, 47}
\definecolor{darkBlue}{RGB}{31, 53, 125}
\definecolor{brightYellow}{RGB}{245, 210, 10}   
\title{\text{Additive drift with tail bounds}}

@@

\end{proof}
\subsection{Theorem for sub-gaussian processes}
\cm
\subsection{Comparing requirements}
\label{sec:compare}
% To use some versions of drift theorems, we need some requirements on the sequence of random variables to be satisfied, such as the following:
% $$
%     \exists \delta \geq 0,\ c \in \mathbb{R} \ \forall  t \geq 0 \ \forall \gamma \in [0, \delta] $$$$
%     E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq e^{\frac{c \gamma^2}{2}},$$

% оr

% $$\exists \delta \geq 0,\ p \geq 1 \ \forall t \geq 0 $$$$
%     E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq 1 - \frac{1}{p}\ .$$

The theorem for sub-Gaussian processes is applicable, when there exist $\delta \geq 0$ and $c \in \mathbb{R}$ such that for all $t \geq 0$ and for all $\gamma \in [0, \delta]$ we have

$$  E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq e^{\frac{c \gamma^2}{2}}.$$

At the same time the negative drift theorem is applicable when there exist $\delta \geq 0$ and $p \geq 1$ such that for all $t \geq 0$ we have

$$
    E\left[e^{\gamma (X_{t + 1} - X_{t})}\right] \leq 1 - \frac{1}{p}\ .$$

To compare rigors of requirements we need to explore behavior $E[e^{\gamma X}]$ depending to $\gamma$. This behavior depends on the distribution $X$, hence we consider the two cases for continuos and for discrete distributions.

@@

By the definition of expectation of a function we have
\[
    E[e^{\gamma X}] = \int_{-\infty}^{+\infty} f(x) e^{\gamma x} \,dx,
\]

where $f(x)$ is a probability density function for $X$.

@@

We compute the first derivative of this expectation over $\gamma$

@@

Note that for all $x \leq 0$ we have $e^{\gamma x} \leq 1$ and for all $x \geq 0$ we have $e^{\gamma x} \geq 1 + \gamma x$. Hence since $f(x) \geq 0$ for all x, we have

@@

Therefore,

@@

We denote $c = \int_{0}^{+\infty} x^2 f(x) \, dx \geq 0$.If we integrate the inequality above, we obtain a lower bound on the expectation.

@@

In the same way we consider the expectation of function by definition

\begin{align*}
    E[e^{\gamma X}] = \sum_{i = 1}^{+\infty} e^{\gamma x_i} \Pr[x = x_i].
\end{align*}

@@

When we take the first derivative we can split to two group, first of which are terms with indices from $P$ set and others
% В таком случае, аналогично возьмем производную и рассмотрим пот отдельности слагаемые из множества P и остальные:

@@

We denote $c = \sum_{i \in P} x_i^2 Pr[x = x_i] \geq 0$ and obtain a bound similar to the one for conditions random variables.

@@

Since the bounds for both cases are the same, we do not distinguish cases for continuos and discrete v.v.

@@

In this case, the expectation of the exponent cannot be lower than 1, and also in the some neighborhood of zero it grows faster than any $e^{\frac{\gamma^2 c}{2}}$, since

@@

Note that $g(0) = 0$, while $E[X] > 0$, therefore, that in any neighborhood of zero we have

@@

Hence, both requirements are not feasible.

@@

In this section we consider some specific family of continuos distributions with symmetric probability density functions, because their expectation equals 0.

@@

\begin{align*}
    \infint{f(x)} & = \infint{c\expx{a}} = 1,
\end{align*}

@@

We denote

@@

Then, we have

@@

We compute the n-th momentum of $X$ (expectation от $X^n$) as follows

@@

\begin{align*}
    E[e^{\gamma X}] = E\left[\sum_{n = 0}^{+\infty}\frac{(\gamma X)^n}{n!}\right] & = \sum_{n = 0}^{+\infty}\gamma^n\frac{E[X^n]}{n!} = \sum_{n = 0}^{+\infty}\gamma^{2n}\frac{E[X^{2n}]}{2n!} =                                \\
    & = \sum_{n = 0}^{+\infty}\frac{(\gamma^2)^{n}}{n!}\left(\frac{E[X^{2n}]n!}{2n!}\right) = \sum_{n = 0}^{+\infty}\frac{(\gamma^2)^{n}}{n!} a_n,
\end{align*}

Where by $a_n$ we denote $\frac{E[X^{2n}]n!}{2n!}$ for all $n \in \mathbb{N}$.

@@

Consider the limit of the sequence $\{\sqrt[n]{a_n}\}_{n \in \mathbb{N}}$

@@

We now consider two cases depending on the value of $a$
\begin{caseof}
    \case{$a \geq 2$}{

        Then $A$ exist, which implies that there exist $C > 0$ such that
        $\forall n \geq 0 a_n \leq C^n$.
        Hence $E[e^{\gamma X}] \leq e^{\gamma^2 C}.$
    }
    \case{$a < 2$}{

@@

We can compute same limit for $\{\sqrt[n]{\frac{a_n}{n!}}\}_{n \in \mathbb{N}}$ and threshold regarding to $a$ is 1, where only for $a \geq 1$ series converge.

@@

For $a \in (1, 2)$ the asymptotic is greater that square root. Hence sub-gausality is not feasible.

@@

Hence both requirements are feasible, since we have

@@

Consequently proving that the process is sub-gaussian is tantamount to finding $p$ in the second requirements.

@@

If $E[X] > 0$ then in some neighborhood around $\gamma = 0$ would be greater than any $e^{\frac{c \gamma^2}{2}}$. To proof it, consider function $G_c(\gamma) = e^{\frac{c \gamma^2}{2}}$, then

$$\frac{\partial G_c(\gamma)}{\partial \gamma} = c\gamma\ G_c(\gamma) = g(\gamma),$$

@@

Example for $X \sim \mathcal{N}(1, 1)$:

\begin{figure}[h!]
    \begin{center}
     \begin{tikzpicture}[scale = 0.8]
        \begin{axis}[
            legend style=
            {nodes={scale=0.9, transform shape},
            at={(1.33,1)},anchor=north east},
            axis lines = left,
            xlabel = \(\gamma\),
            ylabel = {\(f(\gamma)\)},
            xmin=0, xmax=0.4,
            ymin=0.95, ymax=1.5
        ]

@@

        \addplot[color=violet,
            domain=0:3,
            dashed
        ]{1};

        \addlegendentry{\(G_0(\gamma)\)}

        \addplot[color=purple,
            domain=0:0.6,
            line width=0.8pt,
            dashed
        ]{exp((4 * x^2) / 2)};

        \addlegendentry{\(G_4(\gamma)\)}

        \addlegendentry{\(G_{10}(\gamma)\)}

        \addplot[color=orange,
            domain=0:0.6,
            line width=0.8pt,
            dashed
        ]{exp((20 * x^2) / 2)};

        \addlegendentry{\(G_{20}(\gamma)\)}

        \addplot[color=brightYellow,
            domain=0:0.3,
            line width=0.8pt,
            dashed
        ]{exp((50 * x^2) / 2)};

        \addlegendentry{\(G_{50}(\gamma)\)}

        \addplot[color=brightYellow,mark=*] coordinates {(0.0408, 1.0425)};
        \addplot[color=orange,mark=*] coordinates {(0.1053, 1.1172)};
        \addplot[color=red,mark=*] coordinates {(0.2222, 1.2801)};

        \end{axis}
     \end{tikzpicture}
    \end{center}
\end{figure}

\subsection*{Conclusions}

In the case of $\textbf{\text{zero}}$ expectation we cannot always guarantee sub-gausality, which has been proved by the example of series of even functions $\expx{a}$.
